---
title: "SFUSD Multilingual Learners"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      include = TRUE,
                      message = FALSE,
                      fig.width = 8)

pacman::p_load(tidyverse, janitor, readxl, writexl, here, plotly, cowplot, geomtextpath)
percentage <- function(x){
  paste(round(100*x, 2), "%", sep="")
}

mytable1 <- function(x){
  addmargins(table(x, exclude = NULL, deparse.level = 0))
}

mytable2 <- function(x, y){
  addmargins(table(x, y, exclude = NULL, deparse.level = 0))
}

checkdup <- function(df){
  df %>% group_by(student_no) %>% filter(n() > 1) %>% ungroup() %>% arrange(student_no)
}

checkna <- function(df){
  sapply(df, function(x){sum(is.na(x))})
}
```


```{r}
# if (!require('devtools')) install.packages('devtools')
# devtools::install_github('rstudio/leaflet')
```


```{r, layout="l-screen-inset shaded", include = TRUE}

```

```{css, echo=FALSE}
/* Whole document: */
body{
  font-family: Helvetica;
  font-size: 14pt;
}
```


```{r}
ml <- read_xlsx(here("./data/ml.xlsx")) %>%
  mutate(elas = factor(elas,
                       levels = c("English or American Sign Language Only", 
                                  "Initial Fluent English Proficient", 
                                  "Reclassified Fluent English Proficient",
                                  "English Learner", 
                                  "Adult English Learner"),
                       labels = c("EO",
                                  "IFEP", 
                                  "RFEP", 
                                  "EL", 
                                  "EL")),
         grade = factor(grade,
                        levels = c("PK", "TK", "K", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "5th YR Senior"),
                        labels = c("PK-K", "PK-K", "PK-K", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10", "11", "12", "12")),
         Cohort = case_when(grepl("HS", sch) ~ "HS",
                            grepl("MS", sch) ~ "MS",
                            grepl("K-8", sch) ~ "K-8",
                            grepl(" ES", sch) ~ "ES",
                            grepl("EES|Pre-K|PreK|EED|Prek", sch) ~ "PreES")) %>% 
  # foo <- ml %>% select(Cohort, sch) %>% unique()
  drop_na() %>% 
  rename(School = sch,
         ELAS = elas,
         Grade = grade,
         Typology = typo)
# mytable1(ml$Grade)

# EL only
mlml <- ml %>% 
  filter(ELAS == "EL") %>% 
  mutate(Typology = factor(Typology,
                       levels = c("EL 6+ Yrs LTEL", 
                                  "EL 4-5 Yrs Potential LTEL", 
                                  "EL 4+ Yrs On Track", 
                                  "EL 0-3 Yrs", 
                                  "EL 0-3 Yrs Newcomer"),
                       labels = c("LTEL", 
                                  "PLTEL", 
                                  "ELOT", 
                                  "NewEL", 
                                  "Newcomer")))
```


At SFUSD, more than half of our students live in households with a home language other than English, which include English Learner students (ELs), Re-designated Fluent English Proficient students (RFEPs), and Initial Fluent English Proficient students (IFEPs)^[Journey from EL to RFEP: https://drive.google.com/file/d/1XUxLCgTnHKA_DfFEQrCGvngFmXooIBx4/view]. In contrast, English or American Sign Language Only students (EOs) are less than half.

As of `r Sys.Date()`^[Data source: SFUSD Synergy data system], there are a total of `r nrow(mlml)` ELs actively enrolled in SFUSD, making up to `r percentage(nrow(mlml)/nrow(ml))` of the student population^[Student population: the total of students in SFUSD schools excluding Charter schools and Central Enrollment students who have yet enroll in a particular school]. Below is a breakdown of the proportions of four major English Language Acquisition Status, or ELAS: EL, RFEP, IFEP, and EO.

```{r fig.height = 2, fig.width = 8}
ml %>%
  mutate(x = "SFUSD") %>%
  group_by(ELAS) %>% 
  mutate(Percentage = round(n()/nrow(ml), 4) * 100) %>% 
  ungroup() %>% 
  select(x, Percentage, ELAS) %>% 
  unique() %>% 
  ggplot(aes(x, Percentage, fill = ELAS)) +
  geom_col(width = 0.4) +
  geom_text(aes(label = paste0(Percentage, "%")),
            position = position_stack(vjust = 0.5),
            alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  theme_bw(base_size = 12) +
  labs(fill = "ELAS:  ") +
  theme(legend.position = "bottom",
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank()) +
  guides(fill = guide_legend(title = "", nrow = 2, byrow = TRUE)) +
  coord_flip() +
  guides(fill = guide_legend(reverse=TRUE))
```


Let's take a closer look at ELAS enrollment at each of the five cohorts: high schools, middle schools, K-8 schools, elementary schools, and pre elementary schools (early elementary, pre-K, etc.).

```{r fig.width = 8, fig.height = 8}
ml %>%
  group_by(Cohort) %>% 
  mutate(n1 = n()) %>% 
  ungroup() %>% 
  group_by(Cohort, ELAS) %>% 
  mutate(n2 = n(),
         Percentage = round(n2/n1, 4) * 100) %>% 
  ungroup() %>% 
  select(Cohort, Percentage, ELAS) %>% 
  unique() %>% 
  ggplot(aes(Cohort, Percentage, fill = ELAS)) +
  geom_col() +
  geom_text(aes(label = paste0(Percentage, "%")),
            size = 4,
            alpha = 0.5,
            position = position_stack(vjust = 0.6)) +
  geom_vline(xintercept = 1:13 - 0.5, color = "grey90") +
  geom_hline(yintercept = 0:3 * 5, color = "grey90") +
  scale_fill_manual(values = c("steelblue", "dodgerblue4", "darkorange", "yellow"))+
  theme_minimal() +
  theme(legend.position = "bottom",
        axis.title.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.text.x=element_text(face=1, color="grey40", hjust=0, size=14),
        axis.title.y=element_blank(),
        axis.text.y=element_blank()) +
  guides(fill = guide_legend(title = "ELAS: ", nrow = 1, byrow = TRUE)) +
  coord_curvedpolar() 
```


The following five plots provide a closer look at the up-to-date student enrollment by ELAS at each school. Each plot shows schools in each of the five cohorts. 

Note that in each plot, each bar consists of the proportions of four ELAS groups within each school. **Also note that the bars are ordered by the proportion of ELs (which is the first proportion in the bar) from the highest to the lowest**. Do not rely on the color theme, hove over a specific proportion of the colored bar to find out the specific ELAS group and student counts.


```{r}
plot_elas <- function(df){
  p <- df %>% 
    left_join(
      df %>% 
        group_by(School) %>%
        mutate(share = Enrollment/sum(Enrollment)) %>% 
        ungroup() %>% 
        filter(ELAS == "EL") %>% 
        select(School, share)
      ) %>% 
    mutate(share = case_when(is.na(share) ~ 0,
                             .default = share),
           School = fct_reorder(School, share)) %>% 
    ggplot(aes(School, Enrollment, fill = ELAS)) + 
    geom_col(position = "fill") +
    #geom_text(aes(label = scales::percent(share))) +
    scale_fill_brewer(palette = "Set2") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw(base_size = 12) +
    theme(axis.title.y=element_blank(),
          axis.title.x=element_blank(),
          legend.position="none") +
    coord_flip() 
  ggplotly(p)
}

plot_elas_1 <- function(df){
  p <- df %>% 
    left_join(
      df %>% 
        group_by(School) %>%
        mutate(share = Enrollment/sum(Enrollment)) %>% 
        ungroup() %>% 
        filter(ELAS == "EL") %>% 
        select(School, share)
      ) %>% 
    mutate(share = case_when(is.na(share) ~ 0,
                             .default = share),
           School = fct_reorder(School, share)) %>% 
    ggplot(aes(School, Enrollment, fill = ELAS)) + 
    geom_col(position = "fill") +
    #geom_text(aes(label = scales::percent(share))) +
    scale_fill_brewer(palette = "Set3") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw(base_size = 12) +
    theme(axis.title.y=element_blank(),
          axis.title.x=element_blank(),
          legend.position="none") +
    coord_flip() 
  ggplotly(p)
}
```


```{r}
ml %>%
  filter(Cohort == "HS") %>%
  group_by(School, ELAS) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, ELAS, Enrollment) %>% 
  unique() %>% 
  plot_elas()
```


```{r}
ml %>%
  filter(Cohort == "MS") %>%
  group_by(School, ELAS) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, ELAS, Enrollment) %>% 
  unique() %>% 
  plot_elas()
```



```{r fig.height = 3.5}
ml %>%
  filter(Cohort == "K-8") %>%
  group_by(School, ELAS) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, ELAS, Enrollment) %>% 
  unique() %>% 
  plot_elas()
```


```{r fig.height = 12, fig.width = 9}
ml %>%
  filter(Cohort == "ES") %>%
  group_by(School, ELAS) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, ELAS, Enrollment) %>% 
  unique() %>% 
  plot_elas()
```


```{r fig.height = 3.5}
ml %>%
  filter(Cohort == "PreES") %>%
  group_by(School, ELAS) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, ELAS, Enrollment) %>% 
  unique() %>% 
  plot_elas_1()
```


# Multilingual Learner Typology (EL Typology)

EL population is diverse in terms of where they are on the spectrum toward English language proficiency. EL typology is designed to identify different subgroups of ELs so that we can provide research-based multilingual programs and interventions that meet their unique learning needs. 

Please refer to [this slide](https://docs.google.com/presentation/d/12GjkZb5EjsPEqNjqSC97w08JtjseHqMAqfdzIRYsx1E/edit#slide=id.g288282fe4fd_0_4) to understand more about how SFUSD identify different typologies of ELs. Here is a brief glossary of EL Typology categories being used in this website:

 - Newcomer: EL born outside of the US who has been enrolled as an EL for 3 years or less
 - NewEL: EL born in the US who has been enrolled as an EL for 3 years or less
 - ELOT: EL being enrolled as an EL for 4 or more years, but NOT PTEL or LTEL
 - PTEL: EL in grades 3-12 who has been enrolled as an EL for 4-5 years but has not achieved English proficiency level nor ELA grade-specific standard in State tests 
 - LTEL: EL being enrolled as an EL for 6 or more years without making progress 

Below is a breakdown of the proportions of the five EL Typology groups at SFUSD:

```{r fig.height = 2, fig.width = 8}
mlml %>%
  mutate(x = "SFUSD") %>%
  group_by(Typology) %>% 
  mutate(Percentage = round(n()/nrow(mlml), 4) * 100) %>% 
  ungroup() %>% 
  select(x, Percentage, Typology) %>% 
  unique() %>% 
  ggplot(aes(x, Percentage, fill = Typology)) +
  geom_col(width = 0.4, alpha = 0.7) +
  geom_text(aes(label = paste0(Percentage, "%")),
            position = position_stack(vjust = 0.5),
            alpha = 0.7) +
  scale_fill_brewer(palette = "Set1") +
  theme_bw(base_size = 12) +
  labs(fill = "EL Typology: ") +
  theme(plot.title = element_text(size = 14),
        plot.subtitle = element_text(size = 10),
        legend.position = "bottom",
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank()) +
  guides(fill = guide_legend(title = "", nrow = 1, byrow = TRUE)) +
  coord_flip() +
  guides(fill = guide_legend(reverse=TRUE))
```

Certain EL Typologies are more representative than others at a certain grade level, indicating that multilingual pathway programs and interventions need to be customized by grade levels to better serve ELs' learning needs. Looking further into EL Typologies by grade level:

```{r fig.height = 7, fig.width = 9}
mlml %>%
  group_by(Grade) %>% 
  mutate(n1 = n()) %>% 
  ungroup() %>% 
  group_by(Grade, Typology) %>% 
  mutate(n2 = n(),
         Percentage = round(n2/n1, 4) * 100) %>% 
  ungroup() %>% 
  select(Grade, Percentage, Typology) %>% 
  unique() %>% 
  ggplot(aes(Grade, Percentage, fill = Typology)) +
  geom_col() +
  geom_text(aes(label = paste0(Percentage, "%"), 
                vjust=rep(c(0.5, -0.5), length.out=length(Percentage))),
            alpha = 0.5,
            position = position_stack(vjust = 0.4)) +
  scale_fill_brewer(palette = "Paired")  +
  theme_bw(base_size = 12) +
  labs(fill = "EL Typology: ") +
  theme(plot.title = element_text(size = 14),
        plot.subtitle = element_text(size = 10),
        legend.position = "bottom",
        axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        axis.title.y=element_blank()) +
  guides(fill = guide_legend(title = "", nrow = 1, byrow = TRUE)) +
  coord_flip() +
  guides(fill = guide_legend(reverse=TRUE)) 
```


```{r}
# reorder typology plot by LTEL
plot_typo_1 <- function(df){
  p <- df %>% 
    left_join(
      df %>% 
        group_by(School) %>%
        mutate(share = Enrollment/sum(Enrollment)) %>% 
        ungroup() %>% 
        filter(Typology == "LTEL") %>% 
        select(School, share)
      ) %>% 
    mutate(share = case_when(is.na(share) ~ 0,
                             .default = share),
           School = fct_reorder(School, share)) %>% 
    ggplot(aes(School, Enrollment, fill = Typology)) + 
    geom_col(position = "fill", alpha = 0.7) +
    #geom_text(aes(label = scales::percent(share))) +
    scale_fill_brewer(palette = "Set1") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw(base_size = 12) +
    theme(axis.title.y=element_blank(),
          axis.title.x=element_blank(),
          legend.position="none") +
    coord_flip() 
  ggplotly(p)
}

# reorder typology plot by Newcomer
plot_typo_2 <- function(df){
  p <- df %>% 
    left_join(
      df %>% 
        group_by(School) %>%
        mutate(share = Enrollment/sum(Enrollment)) %>% 
        ungroup() %>% 
        filter(Typology == "Newcomer") %>% 
        select(School, share)
      ) %>% 
    mutate(share = case_when(is.na(share) ~ 0,
                             .default = share),
           School = fct_reorder(School, share)) %>% 
    ggplot(aes(School, Enrollment, fill = Typology)) + 
    geom_col(position = "fill") +
    #geom_text(aes(label = scales::percent(share))) +
    scale_fill_brewer(palette = "Accent") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw(base_size = 12) +
    theme(axis.title.y=element_blank(),
          axis.title.x=element_blank(),
          legend.position="none") +
    coord_flip() 
  ggplotly(p)
}

# reorder typology plot by Newcomer
plot_typo_3 <- function(df){
  p <- df %>% 
    left_join(
      df %>% 
        group_by(School) %>%
        mutate(share = Enrollment/sum(Enrollment)) %>% 
        ungroup() %>% 
        filter(Typology == "Newcomer") %>% 
        select(School, share)
      ) %>% 
    mutate(share = case_when(is.na(share) ~ 0,
                             .default = share),
           School = fct_reorder(School, share)) %>% 
    ggplot(aes(School, Enrollment, fill = Typology)) + 
    geom_col(position = "fill") +
    #geom_text(aes(label = scales::percent(share))) +
    scale_fill_brewer(palette = "Set3") +
    scale_y_continuous(labels = scales::percent) +
    theme_bw(base_size = 12) +
    theme(axis.title.y=element_blank(),
          axis.title.x=element_blank(),
          legend.position="none") +
    coord_flip() 
  ggplotly(p)
}
```


Similar to the previous section, below five plots provide a closer look at the up-to-date student enrollment by EL Typology at each school. 

Note that for the following three cohorts, high schools, middle schools, and K-8, **the bars are ordered by the proportion of LTELs (which is the last proportion in the bar) from the highest to the lowest**. Do not rely on the color theme, hove over a specific proportion of the colored bar to find out the specific ELAS group and student counts.

Note that for the following three cohorts, high schools, middle schools, and K-8, the bars are ordered by the first proportion - the proportion of LETLs from the highest to the lowest:

```{r}
mlml %>%
  filter(Cohort == "HS") %>%
  group_by(School, Typology) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, Typology, Enrollment) %>% 
  unique() %>% 
  plot_typo_1()
```


```{r}
mlml %>%
  filter(Cohort == "MS") %>%
  group_by(School, Typology) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, Typology, Enrollment) %>% 
  unique() %>% 
  plot_typo_1()
```


```{r fig.height = 3.5}
mlml %>%
  filter(Cohort == "K-8") %>%
  group_by(School, Typology) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, Typology, Enrollment) %>% 
  unique() %>% 
  plot_typo_1()
```


For the following two cohorts, elementary and pre-elementary schools, the bars are ordered by the first proportion - the proportion of Newcomers from the highest to the lowest:

```{r fig.height = 12, fig.width = 9}
mlml %>%
  filter(Cohort == "ES") %>%
  group_by(School, Typology) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, Typology, Enrollment) %>% 
  unique() %>% 
  plot_typo_2()
```


```{r fig.height = 3.5}
mlml %>%
  filter(Cohort == "PreES") %>%
  group_by(School, Typology) %>%
  mutate(Enrollment = n()) %>% 
  ungroup() %>% 
  select(School, Typology, Enrollment) %>% 
  unique() %>% 
  plot_typo_3()
```


