---
title: "Welcome to MPD Dashboard"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,
                      warning = FALSE,
                      R.options = list(width = 90),
                      include = FALSE)
```


```{r}
# if (!require('devtools')) install.packages('devtools')
# devtools::install_github('rstudio/leaflet')
```


```{r, layout="l-screen-inset shaded", include = TRUE}
# layout = "1-page"
```

```{css, echo=FALSE}
/* Whole document: */
body{
  font-family: Helvetica;
  font-size: 12pt;
}
```

Welcome to MPD Dashboard: the data home of Multilingual Learners at SFUSD.

In the United States, English Learners (ELs) are a growing body of K-12 student population. Under the *Every Student Succeeds Act*, states must annually assess the English language proficiency of ELs and have accountability systems in position to ensure the equitable education outcomes for ELs. 

At SFUSD, *An asset-based mindset towards MLs* is the super power of SFUSD educators and we refer to ELs as **Multilingual Learners** (MLs) because we believe that multilingualism and multiliteracy is the outcome of our education for this unique student group.

**"Serving Multilingual Learners is a central responsibility of each and every educator. Effectively serving this diverse group of learners fosters excellence for all students because language is foundational to learning and development in all students." ([SFUSD Roadmap for Multilingual Learner Achievement and Success](https://drive.google.com/file/d/1U6bHk_8so1Egx9f5CF8izHgYJlp-MzTG/view))**



